\newpage
\section*{Problems}

\begin{multicols}{2}
  \begin{enumerate}[itemsep=4pt,leftmargin=12pt]
%  %  \question Distinguish between diffraction and refraction.
%  %  \vspace{1.2in}
%
%  \question An ocean wave diffracts as it travels through an opening in a
%  barrier in a harbour. Which of the following variables can be altered to
%  change the amount of diffraction the wave experiences? (The amount of
%  diffraction refers to the spacing between fringes.)
%  \begin{enumerate}[noitemsep,topsep=0pt,label={\Roman*.}]
%  \item distance between the source of the waves and the barrier
%  \item amplitude
%  \item frequency
%  \item wavelength
%  \item slit width
%  \end{enumerate}
%  \begin{oneparchoices}
%    \choice I and II
%    \choice II, III, and IV
%    \choice I, III, and IV
%    \choice III, IV, and V
%    \choice I and IV
%  \end{oneparchoices}
%  
%  \question As a student is performing a double slit experiment to determine the
%  wavelength of a light source, she realizes that the nodal lines are too close
%  together to be accurately measured. To increase the distance between the
%  nodal lines, she could:
%  \begin{choices}
%    \choice decrease the slit separation
%    \choice decrease the distance between the slits and the screen
%    \choice move the light source closer to the slits
%    \choice both A and B
%    \choice A, B and C
%  \end{choices}
%
%  \question An observer can hear sound from around a corner but cannot see light
%  from around the same corner. Which of the following helps to explain
%  this phenomenon?
%  \begin{choices}
%    \choice Sound is a longitudinal wave, and light is an electromagnetic wave.
%    \choice Sound is a mechanical wave, and light is a transverse wave.
%    \choice Light travels at a speed much faster than that of sound.
%    \choice Light has a much shorter wavelength than sound.
%  \end{choices}
%  
%  \question The brightest and clearest diffraction patterns are seen when light
%  is shone through:
%  \begin{choices}
%    \choice a single slit
%    \choice a double slit
%    \choice a diffraction grating
%    \choice a polarizing filter
%    \choice a thin soap film
%  \end{choices}
%  
%  \question The greater the number of lines on a diffraction grating of a
%  given size,
%  \begin{choices}
%    \choice the greater the range of wavelengths that can be diffracted
%    \choice the smaller the range of wavelengths that can be diffracted
%    \choice the greater the distance between the bright fringes produced
%    \choice the smaller the distance between the bright fringes produced
%    \choice the less clear the diffraction pattern will be
%  \end{choices}
%  \newpage
%  
%  %\question What is dispersion, and how is it related to diffraction and
%  %refraction?

  \item A prism bends blue light more than red. Is the same true of a
    diffraction grating? Explain using relevant equations.
  
  \item Upon observing an interference pattern produced onto a screen, how
    could you identify whether a single slit or double slit produced the
    pattern?
    
  \item Blue light ($\lambda=\SI{475}{\nano\metre}$) is sent through a single
    slit with a width of \SI{2.1}{\micro\metre}. What is the maximum possible
    number of bright fringes produced on the screen?

  \item Determine the distance that the \emph{third} bright fringe would lie
    from the central diffraction maximum in a single slit diffraction pattern
    generated with \SI{542}{\nano\metre} light incident on a \SI{1.2e-4}{\metre}
    slit falling onto a screen \SI{68}{\centi\metre} away.

  \item A beam of purple light, which consists of red
    ($\lambda_r=\SI{702}{\nano\metre}$) and violet
    ($\lambda_v=\SI{404}{\nano\metre}$) light, is sent through a single-slit.
    Predict which colour will have a wider central maximum in the diffraction
    pattern. Calculate the difference if the purple light is incident on a
    \SI{6.9e-5}{\metre} wide slit falling onto a screen \SI{85}{\centi\metre}
    away.

  \item A double slit apparatus is held \SI{1.2}{\metre} from a screen. When red
    light ($\lambda=\SI{600}{\nano\metre}$) is sent through the double slit, the
    interference pattern on the screen shows a distance of
    \SI{12.5}{\centi\metre} between the first and tenth dark fringes. What is
    the separation of the slits?
    %\part What will be the difference in path length for the waves travelling
    %from each slit to the tenth nodal line?
  
  \item The signal from a \SI{103.9}{\mega\hertz} FM radio station reflects off
    a building \SI{400}{\metre} away, effectively creating two sources of the
    same signal. You are driving at \SI{60}{\kilo\metre\per\hour} along a road
    parallel to a line between the station's antenna and the building, and
    located at a perpendicular distance of \SI{6.5}{\kilo\metre} from them. How
    often does the signal appear to fade when you are driving along the road?
    The speed of radio waves is \SI{3.00e8}{\metre\per\second}. Please answer to
    \emph{three} significant figures.

  \item A beam of parallel rays from a \SI{29}{\mega\hertz} radio transmission
    passes between two electrically conducting (therefore opaque to radio waves)
    buildings \SI{45}{\metre} apart. What is the angular width (i.e.\ what is
    the angle $\theta$ of the first minima) of the beam when it emerges from
    between the buildings?
  
  \item A skydiver is falling towards the ground. How close to the ground will
    she have to be before she is able to distinguish two yellow baseballs lying
    \SI{25.0}{\centi\metre} apart, reflecting \SI{570}{\nano\metre} light in
    air? Her pupil diameter is \SI{3.35}{\milli\metre}. The speed of light
    inside the human eye is \SI{2.21e8}{\metre\per\second}.

  \item The newest reconnaissance (i.e.\ spy) satellites can resolve objects
    separated by only \SI{3.0}{\centi\metre}. If one of these satellites orbits
    Earth at low orbital altitude of \SI{250}{\kilo\metre}, determine the size
    of the satellites' circular imaging aperture. (Use \SI{455}{\nano\metre}
    light for the light in the lenses of the satellites.)
  %  \part Describe why the value from part (a) is a theoretical best-case
  %  result. What other effects would play a role in a satellite’s ability to
  %  resolve objects on the surface of Earth?
  \end{enumerate}
\end{multicols}
